Molecular Gas Dynamics of the lensed wet-merger RXSJ1131-1231 at $z$=0.654

### 2016-06-29 WRITEUP/ms.tex
minor typos

In this paper, we explore the ISM properties of the quadruply imaged
quasar RXS J113151.62-123158 (hereafter RXJ1131) at
$z_\textrm{AGN}$\,$=$\,0.685, with an Einstein ring of size
1\farcs83 in radius. The foreground lensing
galaxy is an elliptical galaxy at $z_\textrm{L}$\,$=$\,0.295. The redshifts
are spectroscopically confirmed by \citet{Sluse03a}.
A black hole mass estimate of M$_{\rm BH}$\,$<$\,2\E{8}\Msun
is also reported based on X-ray observations \citep{Reis14a}.
\defcitealias{Sluse03a}{S03}

@@

\subsubsection{Interpretation of the Source-plane Morphology} \label{sec:caveat}
The reconstructed source locations in \Fig{model} demonstrate
an intrinsic velocity gradient across the source plane, which is
indicative of a kinematically-ordered disk-like galaxy.
Additional support to the disk conjecture
can be found in the double-horned line profile (\Fig{CO21spec})
and the observed (image plane) velocity field (\Fig{CO21highO}). Furthermore,
\citetalias{Claeskens06a} also find that the reconstructed source plane emission in optical-NIR
is best-reproduced using a $n$\,=\,1 Sersic profile.
We thus interpret RXJ1131 to be a disk galaxy.

@@

\subsubsection{Linewidth and Sizes} \label{sec:sizes}
A linewidth of $\Delta v_{\rm FWHM}$$\sim$600\pmm160\,\kms found from
fitting a single Gaussian
is considerably larger that those in the \citepalias{Combes13a} sample
(370\,\kms) as well as local ULIRGs
\citep[300\pmm85\,\kms, with the largest being 480\,\kms;][hereafter S97]{Solomon97a}.
%ULIRGs (z < 0.3) studied in CO(1-0) by Solomon97a: ≥ 1/3 of them have double-peaked or flat-topped spectrum. But their linewidths are based on single Gauss fit.
Yet, given the dynamic nature of these galaxies,
a CO linewidth of $\sim$600 may not be surprising.
Indeed, a linewidth of $\Delta v$ = 750\,\kms has been observed
in a local LIRG (Arp 118) \citep{SV05a}.
Such FWHM is also commonly observed in high-$z$ starburst galaxies
\citep[780\pmm320\,\kms; \eg][]{Greve05a}.
\defcitealias{Solomon97a}{S97}\defcitealias{Greve05a}{G05}\defcitealias{SV05a}{SV05}

@@

\subsubsection{Gas Mass and Gas Ratios}
% Solomon97 (ULIRGs z < 0.3): 0.4-1.5 e10 Msun
% SV05: Solomon+97: average Mgas ~7e9 Msun
% SV05: high-z: (2.5-10)e10 Msun
A total molecular gas mass of
$M_{\rm gas}^{\rm total}$ = $($1.57\pmm0.38$)$\E{10}\,\Msun is
residing within the merger system (see \Sec{caveat}).
This is comparable to those found in the
\citetalias{Combes13a} sample %(1.5\E{10}\Msun)
as well as local U/LIRGs
\citep[][\citetalias{Solomon97a}]{Sanders96a} but higher than
in local LIRGs \citepalias{Ueda14a}, as expected.
Our finding is thus consistent with previous results, where
high-$z$ galaxies tend to have higher molecular gas mass %\footnote{Assuming the same conversion factor}
and is
distributed over larger spatial extents
\citepalias[\eg][]{SV05a,Riechers11a,Ivison11a}.
% gas-to-dyn ratio
% DS98: f ~ 0.16 (but w/in disc or ring of R ~ 0.5 kpc)
% Bothwell+13: f ~ 0.3? (but over larger R)
% --> ULIRGs gas fraction is even lower if average over same scale as SMGs
% Daddi10a; Tacconi10a: f_gas ? 40% in z ? 1 discs, significantly higher than local discs (but alphaCO not same as ULIRGs)
Using the dynamical mass and the total
gas mass ($M_{\rm gas}^{\rm total}$), we find a gas mass
fraction of $f_{\rm gas-dyn}$ = 0.19\pmm0.37, which is
consistent with the results reported by \citetalias{Combes13a}:
a gas fraction of 15\% in ULIRGs at 0.2\,$<$\,$z$\,$<$\,0.4 and 24\%
in those at 0.6\,$<$\,$z$\,$<$\,1.0, and those from resolved measurements of local ULIRGs \citepalias{Downes98a}.
% This implies that the gas fraction in ULIRGs at $z$$<$\,1 is
% similar to high-z starburst galaxies, but
% lower if we average over the same spatial scales (as SMGs, which
% are more extended; \citet{Bothwell13a}).
This implies that the gas fraction in ULIRGs at $z$\,$<$\,1 is
similar to those in high-$z$ ``normal'' star-forming disk galaxies
\citep[assuming an \alphaco conversion factor of 0.8 \alphaU;][]{Daddi10a,Tacconi10a} as well as
high-$z$ starburst galaxies
\citep[with similar CO sizes between the two populations (see \Sec{sizes}); cf. smaller CO sizes from earlier studies; \citetalias{Downes98a}]{Bothwell13a}.
\defcitealias{Daddi10a}{D10}

@@

There are a number of systematic uncertainties associated with this quantity, \eg
the mass opacity coefficient $\kappa$,
the \alphaco conversion factor, and the brightness temperature ratio $R_{\rm 21}$.
The fact that RXJ1131 has a lower dust-to-gas ratio 
and appears to be more dusty than most ULIRGs at the same epoch
may imply different physical
environment compared to others. If we instead use the ``Galactic" \alphaco value, 
the gas mass (and thus gas-to-dust ratio) would be $\sim$6 times higher.
On the other hand, we would also obtain a higher gas mass if 
we assume subthermal excitation between \bco and \aco line emission.
However, we expect this to be a minor effect as ULIRGs are observed to be
thermalized up to $J$ = 3 or 4.

@@

To first order, the star formation efficiency
$($SFE = \LFIR$/$$M_{\rm gas})$
indicates the \SF rate per
unit solar mass of molecular gas available in a galaxy.
Using a wavelength range of 40\,$-$\,500\,$\micron$ defined
in \citetalias{Combes13a} as the far-IR luminosity,
we find an SFE of 58\pmm10 \Lsun $M_{\odot}^{-1}$,
which is on the low end among other U/LIRGs
% \citep[$\sim$180\pmm160 \Lsun $M_{\odot}^{-1}$;][]{Solomon97a, Combes11a}
\citep{Solomon97a, Combes11a}
but 
similar to those of
low-$z$ spiral galaxies \citepalias[$z$\,$<$\,0.1;][]{SV05a}.
In constrast, a star formation efficiency of $>$100 \Lsun $M_{\odot}^{-1}$
has been reported in ULIRGs and SMGs at 0.6\,$<$\,$z$\,$<$\,3
\citep[][\citetalias{Riechers11a,Combes13a}]{Tacconi08a}.
\defcitealias{Tacconi08a}{T08}
% Depletion time:
% Greve+05: ~40 Myr
Assuming the \SF continues at the current rate without gas replenishment and is the only
way to consume the gas reservoir, this corresponds to a depletion
time of $\tau$\,=\,102\pmm25\,Myr.
Since the \SF rate is expected to vary in an interacting system and that AGN accretion
also consumes some fraction of the gas, the depletion
timescale should only be considered as an upper limit.

@@

\subsection{Fate of RXJ1131}
The classical picture for mergers is one where they are responsible for the formation of the local 
red and passive spheroidal galaxies.
With more realistic treatments of star formation and feedbacks in recent simulations,
% ? which suppress efficient early star formation ? play a critical role in enabling disc formation in simulations.
it has been suggested that it is possible  to suppress bulge formation 
in gas-rich mergers, thereby evolving into large 
disks \citep{Springel05a, Robertson06a, Hopkins09a}.
% (bulge dominated; Springel05a, Robertson06a)
% and disc with low B/D ratio ($\sim$0.1$-$0.2; Hopkins09a).
%
%In systems that are sufficiently gas-rich, it is possible to 
%suppress the formation of a bulge in mergers of gas-rich discs (even in major merger)
%since there is little/no significant stellar material % (collisionless)
%available to torque away the cold gas in the disc, thus little/no angular momentum is lost, 
%and gas will rapidly reform a cold disc in the background 
%of the relaxing stellar potential.
%Gas, being collisional, cannot violently relax, but must
%have its angular momentum torqued away in order to dissipate and
%build a bulge by forming stars in a central starburst 
%(\ie gas loses angular momentum to stars in the
%perturbed disk and fall toward center).
% In a merger,
%this torque is primarily internal, from stars in the same disc: the
%passages and merger of the secondary induce a non-axisymmetric
%distortion in the primary stellar and gas disc. The stellar distortion
%(in a trailing resonance and close proximity to the gas response)
%efficiently removes angular momentum from the gas and allows it
%to fall to the center and form a bulge  (see also Barnes \& Hernquist
% 1996).
% A bulge will form because cold gas with no angular momentum 
% will be largely unable to form any sort of disk, and that a galaxy?s 
% worth of gas compressed to high densities and small
% radii will inevitably form a large mass in stars). 
%
Since both RXJ1131 and its companion are gas-rich
% unlikely RXJ1131 will evolve into present-day massive 
% ellipticals (M$\gtrsim$1\E{12}\,\Msun), 
% gas-poor mergers: --> present-day massive elliticals
% (Boylan-Kolchin, Ma & Quataert 2005, 2006; Naab, Khochfar & Burkert 2006; Robertson B., Cox T. J., Hernquist L., Franx M., Hopkins P., Martini P., Springel V., 2006a, ApJ, 641, 21; Cox et al. 2006b)
and that the bulk of the angular momentum of the gas is  
retained in minor mergers,
it is likely that RXJ1131 will evolve into either a 
``normal'' elliptical/S0 galaxy or a large disk galaxy.  
% normal E only because gE needs ways to remove angular momentum efficiently for gas to funnel
In the latter, the stellar populations from the companion will be displaced onto the 
dominant galaxy to thicken the disk \citep[\eg][]{Lotz08a, Robertson06a}.
% disk from gas-rich major merger: \citep[\eg][]{Lotz08a, Robertson06a}.
Given that the molecular gas in 
RXJ1131 is distributed in an extended disk configuration and that its 
SFE is on the low end among other ULIRGs,  % not enough stellar material to efficiently remove ang. mom.
it is more conceivable that RXJ1131 will evolve into the
latter.
% Lotz08a: a major gas-rich merger without AGN feedback does not, by itself, produce a red and dead spheroidal galaxy
Other mechanisms e.g. formation of bar-like structures will be needed to remove the
angular momentum of the gas more efficiently to transform the gas disk into a stellar spheroid and
AGN feedbacks are also necessary to halt further activities for RXJ1131 to evolve into an 
E/S0 galaxy \citep{Lotz08a}.

>>>>

In this paper, we explore the ISM properties of the quadruply imaged
quasar RXS J113151.62-123158 (hereafter RXJ1131) at
$z_\textrm{AGN}$\,$=$\,0.685, with an Einstein ring of size
1\farcs83 in radius. The foreground lensing
galaxy is an elliptical galaxy at $z_\textrm{L}$\,$=$\,0.295. The redshifts
are spectroscopically confirmed by \citet[hereafter S03]{Sluse03a}.
A black hole mass estimate of M$_{\rm BH}$\,$<$\,2\E{8}\Msun
is also reported based on X-ray observations \citep{Reis14a}.
\defcitealias{Sluse03a}{S03}

@@

\subsubsection{Interpretation of the Source-plane Morphology} \label{sec:caveat}
The reconstructed source locations in \Fig{model} demonstrate
an intrinsic velocity gradient across the source plane, which is
indicative of a kinematically-ordered disk-like galaxy.
Additional support to the disk conjecture
can be found in the double-horned line profile (\Fig{CO21spec})
and the observed (image plane) velocity field (\Fig{CO21highO}). Furthermore,
\citetalias{Claeskens06a} also find that the reconstructed source plane emission in optical-NIR
is best-reproduced using a $n$\,=\,1 Sersic profile.
We thus interpret RXJ1131 as a disk galaxy.

@@

\subsubsection{Linewidth and Sizes} \label{sec:sizes}
A linewidth of $\Delta v_{\rm FWHM}$$\sim$600\pmm160\,\kms found from
fitting a single Gaussian
is considerably larger that those in the \citepalias{Combes13a} sample
(370\,\kms) as well as local ULIRGs
\citep[300\pmm85\,\kms, with the largest being 480\,\kms;][hereafter S97]{Solomon97a}.
%ULIRGs (z < 0.3) studied in CO(1-0) by Solomon97a: ≥ 1/3 of them have double-peaked or flat-topped spectrum. But their linewidths are based on single Gauss fit.
Yet, given the dynamic nature of these galaxies,
a CO linewidth of $\sim$600 may not be surprising.
Indeed, a linewidth of $\Delta v$ = 750\,\kms has been observed
in a local LIRG (Arp 118) \citep{SV05a}.
Such FWHM is also commonly observed in high-$z$ starburst galaxies
\citep[780\pmm320\,\kms; \eg][hereafter G05]{Greve05a}.
\defcitealias{Solomon97a}{S97}\defcitealias{Greve05a}{G05}\defcitealias{SV05a}{SV05}

@@

\subsubsection{Gas Mass and Gas Ratios}
% Solomon97 (ULIRGs z < 0.3): 0.4-1.5 e10 Msun
% SV05: Solomon+97: average Mgas ~7e9 Msun
% SV05: high-z: (2.5-10)e10 Msun
A total molecular gas mass of
$M_{\rm gas}^{\rm total}$ = $($1.57\pmm0.38$)$\E{10}\,\Msun is
residing within the merger system (see \Sec{caveat}).
This is comparable to those found in the
\citetalias{Combes13a} sample %(1.5\E{10}\Msun)
as well as local U/LIRGs
\citep[][\citetalias{Solomon97a}]{Sanders96a} but higher than
in local LIRGs \citepalias{Ueda14a}, as expected.
Our finding is thus consistent with previous results, where
high-$z$ galaxies tend to have higher molecular gas mass %\footnote{Assuming the same conversion factor}
and is
distributed over larger spatial extents
\citep[\eg \citetalias{SV05a};][]{Riechers11a,Ivison11a}.
% gas-to-dyn ratio
% DS98: f ~ 0.16 (but w/in disc or ring of R ~ 0.5 kpc)
% Bothwell+13: f ~ 0.3? (but over larger R)
% --> ULIRGs gas fraction is even lower if average over same scale as SMGs
% Daddi10a; Tacconi10a: f_gas ? 40% in z ? 1 discs, significantly higher than local discs (but alphaCO not same as ULIRGs)
Using the dynamical mass and the total
gas mass ($M_{\rm gas}^{\rm total}$), we find a gas mass
fraction of $f_{\rm gas-dyn}$ = 0.19\pmm0.37, which is
consistent with the results reported by \citetalias{Combes13a}:
a gas fraction of 15\% in ULIRGs at 0.2\,$<$\,$z$\,$<$\,0.4 and 24\%
in those at 0.6\,$<$\,$z$\,$<$\,1.0, and those from resolved measurements of local ULIRGs \citepalias{Downes98a}.
% This implies that the gas fraction in ULIRGs at $z$$<$\,1 is
% similar to high-z starburst galaxies, but
% lower if we average over the same spatial scales (as SMGs, which
% are more extended; \citet{Bothwell13a}).
This implies that the gas fraction in ULIRGs at $z$\,$<$\,1 is
similar to those in high-$z$ ``normal'' star-forming disk galaxies
\citep[assuming an \alphaco conversion factor of 0.8 \alphaU;][]{Daddi10a,Tacconi10a} as well as
high-$z$ starburst galaxies
\citep[with similar CO sizes between the two populations (see \Sec{sizes}); cf. smaller CO sizes from earlier studies; \citetalias{Downes98a}]{Bothwell13a}.

@@

There are a number of systematic uncertainties associated with this quantity, \eg
the mass opacity coefficient $\kappa$,
the \alphaco conversion factor, and the brightness temperature ratio $r_{\rm 21}$.
The fact that RXJ1131 has a lower dust-to-gas ratio
and appears to be more dusty than most ULIRGs at the same epoch
may imply different physical
environment compared to others. If we instead use the ``Galactic" \alphaco value,
the gas mass (and thus gas-to-dust ratio) would be $\sim$6 times higher.
On the other hand, we would also obtain a higher gas mass if
we assume subthermal excitation between \bco and \aco line emission.
However, we expect this to be a minor effect as ULIRGs are observed to be
thermalized up to $J$ = 3 or 4.

@@

To first order, the star formation efficiency
$($SFE = \LFIR$/$$M_{\rm gas})$
indicates the \SF rate per
unit solar mass of molecular gas available in a galaxy.
Using a wavelength range of 40\,$-$\,500\,$\micron$ defined
in \citetalias{Combes13a} as the far-IR luminosity,
we find an SFE of 58\pmm10 \Lsun $M_{\odot}^{-1}$,
which is on the low end among other U/LIRGs
% \citep[$\sim$180\pmm160 \Lsun $M_{\odot}^{-1}$;][]{Solomon97a, Combes11a}
\citep[\citetalias{Solomon97a};][]{Combes11a}
but
similar to those of
low-$z$ spiral galaxies \citepalias[$z$\,$<$\,0.1;][]{SV05a}.
In constrast, a star formation efficiency of $>$100 \Lsun $M_{\odot}^{-1}$
has been reported in ULIRGs and SMGs at 0.6\,$<$\,$z$\,$<$\,3
\citep[][\citetalias{Combes13a}]{Tacconi08a, Riechers11a}.
\defcitealias{Tacconi08a}{T08}
% Depletion time:
% Greve+05: ~40 Myr
Assuming the \SF continues at the current rate without gas replenishment and is the only
way to consume the gas reservoir, this corresponds to a depletion
time of $\tau$\,=\,102\pmm25\,Myr.
Since the \SF rate is expected to vary in an interacting system and that AGN accretion
also consumes some fraction of the gas, the depletion
timescale should only be considered as an upper limit.

@@

\subsection{Fate of RXJ1131-1231}
The classical picture for mergers is one where they are responsible for the formation of the local
red and passive spheroidal galaxies.
With more realistic treatments of star formation and feedbacks in recent simulations,
% ? which suppress efficient early star formation ? play a critical role in enabling disc formation in simulations.
it has been suggested that it is possible  to suppress bulge formation
in gas-rich mergers, thereby evolving into large
disks \citep{Springel05a, Robertson06a, Hopkins09a}.
% (bulge dominated; Springel05a, Robertson06a)
% and disc with low B/D ratio ($\sim$0.1$-$0.2; Hopkins09a).
%
%In systems that are sufficiently gas-rich, it is possible to
%suppress the formation of a bulge in mergers of gas-rich discs (even in major merger)
%since there is little/no significant stellar material % (collisionless)
%available to torque away the cold gas in the disc, thus little/no angular momentum is lost,
%and gas will rapidly reform a cold disc in the background
%of the relaxing stellar potential.
%Gas, being collisional, cannot violently relax, but must
%have its angular momentum torqued away in order to dissipate and
%build a bulge by forming stars in a central starburst
%(\ie gas loses angular momentum to stars in the
%perturbed disk and fall toward center).
% In a merger,
%this torque is primarily internal, from stars in the same disc: the
%passages and merger of the secondary induce a non-axisymmetric
%distortion in the primary stellar and gas disc. The stellar distortion
%(in a trailing resonance and close proximity to the gas response)
%efficiently removes angular momentum from the gas and allows it
%to fall to the center and form a bulge  (see also Barnes \& Hernquist
% 1996).
% A bulge will form because cold gas with no angular momentum
% will be largely unable to form any sort of disk, and that a galaxy?s
% worth of gas compressed to high densities and small
% radii will inevitably form a large mass in stars).
%
Since both RXJ1131 and its companion are gas-rich
% unlikely RXJ1131 will evolve into present-day massive
% ellipticals (M$\gtrsim$1\E{12}\,\Msun),
% gas-poor mergers: --> present-day massive elliticals
% (Boylan-Kolchin, Ma & Quataert 2005, 2006; Naab, Khochfar & Burkert 2006; Robertson B., Cox T. J., Hernquist L., Franx M., Hopkins P., Martini P., Springel V., 2006a, ApJ, 641, 21; Cox et al. 2006b)
and that the bulk of the angular momentum of the gas is
retained in minor mergers,
it is likely that RXJ1131 will evolve into either a
``normal'' elliptical/S0 galaxy or a large disk galaxy.
% normal E only because gE needs ways to remove angular momentum efficiently for gas to funnel
In the latter, the stellar populations from the companion will be displaced onto the
dominant galaxy to thicken the disk \citep[\eg][]{Lotz08a, Robertson06a}.
% disk from gas-rich major merger: \citep[\eg][]{Lotz08a, Robertson06a}.
Given that the molecular gas in
RXJ1131 is distributed in an extended disk configuration and that its
SFE is on the low end among other ULIRGs,  % not enough stellar material to efficiently remove ang. mom.
it is more conceivable that RXJ1131 will evolve into the
latter.
% Lotz08a: a major gas-rich merger without AGN feedback does not, by itself, produce a red and dead spheroidal galaxy
Other mechanisms e.g. formation of bar-like structures will be needed to remove the
angular momentum of the gas more efficiently to transform the gas disk into a stellar spheroid and
AGN feedbacks are also necessary to halt further activities for RXJ1131 to evolve into an
E/S0 galaxy \citep{Lotz08a}.
